---
title: 'Assignment 1: Data Exploration'
author: "Christian Osadebe"
date: "09-07-2025"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```
## 1. Introduction:
Demonstration of basic data analysis in R using the Cereals dataset. It includes importing data, generating descriptive statistics, transforming variables, and visualizing results.

Data Source: Kent State University. (2025). Cereals.csv [Course dataset]. BA-64060: Fundamentals of Machine Learning. https://kent.instructure.com/courses/120714/pages/csv-files

## 2. Download and Import the Dataset
```{r}
# Load necessary libraries
library(tidyverse)

# Import the dataset
cereals <- read.csv("C:/Users/bondo/Downloads/Cereals.csv", stringsAsFactors = TRUE)

```

## 3. Descriptive statistics of quantitative and categorical variables

```{r}
# Quantitative summary
summary(cereals[, c("calories", "protein", "fat", "carbo", "sugars")])

# Categorical summary
table(cereals$Manufacturer)
table(cereals$Type)


```

## 4. Variable Transformation

```{r}
cereals <- cereals %>%
  mutate(sugar_carb_Ratio = sugars / carbo)


```

## 5. Visualizations

```{r}
library(ggplot2)
# Histogram of Calories
ggplot(cereals, aes(x = calories)) +
  geom_histogram(binwidth = 10, fill = "steelblue", color = "white") +
  labs(title = "Distribution of Calories", x = "calories", y = "count")

# Scatterplot: Sugars vs. Rating
ggplot(cereals, aes(x = sugars, y = rating, color = type)) +
  geom_point() +
  labs(title = "Sugars vs. Rating", x = "Sugars (g)", y = "Consumer Rating")


```







